---
title: Introduction to R for distance sampling
header-includes: \usepackage{color}
author: Centre for Research into Ecological and Environmental Modelling
subtitle: Workshop, 26 August 2018
output: pdf_document
fontsize: 12pt
---

# 1.6 Exporting 'Distance for Windows' projects to R using `readdst`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load library
library(knitr)
library(readdst)

# Specify whether answers are shown 
answer <- TRUE
mycode <- ifelse(answer, TRUE, FALSE)
#answer <- FALSE```
```

The R package `readdst` (Miller, NA) was created to port Distance for Windows projects (Thomas \textit{et al}. (2010) into R to make use of the software developments that continue to be incorporated into R but not always in Distance for Windows. 

In this practical, the ducknests project will be used as an example. The detections from this project were used in a previous practical. 

## Objective of the practical 

1. Install the necessary packages

2. Convert a distance project

3. Examine the resulting object 

4. Extract information from the resulting object.

## Setting up

The package `readdst` is stored on github rather than the CRAN repository, therefore installation is a bit different to that of `Distance` but still straightforward. One wrinkle is that a 32 bit version of R should be used. 

The commands below will install the necessary packages:

`install.packages("devtools")`

`devtools::install_github("distancedevelopment/readdst")`

Once the `readdst` package has been installed, it needs to be loaded into the current work session:

`library(readdst)`

## Converting a project

Converting a project for analysis in R is straightforward:

```{r, echo=T}
# Commands used to extract the data (in R 32bit)
#ducknest <- convert_project("C:/workshops/2018/Ducknest exercise")
#temp <- (ducknest$"Half normal with no adjustments"$env$data)
```

`ducknest <- convert_project("C:/workshops/2018/Ducknest exercise")`

There are two things to note, 

+ the path to the Distance for Windows project should be specified (if not in the working directory)

+ the '.dst' suffix is not required. 

## Examining the R object

The resulting object (in this case `ducknest`) is composed of a series, or list, of objects within objects (a bit like the object from the `ds` function). Here the first level of objects are named after the Analyses in the Distance for Windows project. The `names` function can be used to delve into the objects: 

`names(ducknest)`

There was only one Analysis definition set up for this project and so the command gives:

\color{blue}
`"Half normal with no adjustments"`
\color{black}

Appling `names` to this object shows that there many further objects (note quotes are required):

`names(ducknest$"Half normal with no adjustments")`

\color{blue}
` [1] "call"          "aic.select"    "status"        "env"    `      
` [5] "filter"        "group_size"    "detection_by"  "gof_intervals"`

` [9] "estimation"    "name"          "ID"            "engine" `      
`[13] "project"       "project_file" `
\color{black}

Here, we highlight the `env` object because this object contains the data tables.

`names(ducknest$"Half normal with no adjustments"$env)`

\color{blue}
`[1] "units"        "obs.table"    "sample.table" "region.table" "data"  `
\color{black}

These tables can be extracted from the list (for ease of use) and then analysed using the `ds` function (Miller 2016). For example, 

`duck.detect <- ducknest$"Half normal with no adjustments"$env$data`

`ds(duck.detect, key="hn")`

See practical 2 for an example of including `sample.table` etc as arguments. 

## Further information

More information and examples are provided at the following links. 

A poster outlining the process:

http://converged.yt/posters/RexstadISEC2018poster.pdf

Examples of converting data (Amakihi and golftees):

https://cdn.rawgit.com/DistanceDevelopment/readdst/e308e819/vignettes/Project-migration.html

https://cdn.rawgit.com/DistanceDevelopment/readdst/e308e819/vignettes/golftees.html

## References

Miller DL (NA) `readdst`: Convert Distance for Windows projects to R analyses. R package version 0.0.5.

Miller DL (2016). Distance: Distance Sampling Detection Function and Abundance Estimation. R package version 0.9.6. https://CRAN.R-project.org/package=Distance

Thomas L, Buckland ST, Rexstad EA, Laake J, Strindberg S, Hedley SL, Bishop JR, Marques TA and Burnham KP (2010) Distance software: design and analysis of distance sampling surveys for estimating population size. Journal of Applied Ecology: 47(1):5-14


